Bare Demo of IEEEtran.cls\\ for IEEE Conferences

### 2016-11-23 paper.tex
Definition of ILEs acronym
Interactive Learning Environment, could also be Intelligent but not yet.

When designing instruction, an objective is to assign learning activities
that result challenging to students, but not much as to frustrate them. For
this, experienced instructors gauge a student’s affective state and then assign
an activity with the appropriate level of difficulty. However, recognition of 
emotions is a difficult task, even
for humans. Instructors use their social intelligence often to recognize
students’ affective states. In class a good instructor habitually reads the
faces of students in the classroom to see if they are confused, bored or
engaged; then decides what to do next. ILEs can be enhanced if they can offer an
automatic perception of emotions. The recognition and simulation of human
affects are becoming important fields of study, as many researchers have
demonstrated affect-aware computers can provide better performance in assisting
humans \cite{picard2001toward}. When ILEs embrace these methodologies and techniques a
problem arises. In order to perform recognition of the users' affective states,
sensors must be used to gather data. Some sensors rely on physiological readings
and must be in physical contact with students. These sensors can be considered
intrusive or invasive, and can disrupt the student’s learning experience 
\cite{zhai2008stress,sidney2005integrating,arroyo2009emotion}.
Other sensors such as video cameras can also be considered invasive, since they always transmit the
identity, appearance and behaviour along with emotional data \cite{picard2001toward}.

In this work, a method for the recognition of affective states through keystroke
and mouse dynamics is proposed. The hypothesis is that students vary the dynamic
of keystrokes according to their affective state when programming.  The method
has three main components: a Javascript library to capture keyboard and mouse
dynamics from a browser-based editor, a pre-processing step whose output is a
feature vector that is sent to the third component, a classification algorithm.
An experiment was conducted where students solved a series of programming
exercises using a web based learning environment. Using only the data from the
student’s keyboard and mouse dynamics six affective states where recognized for
each of the attempts at solving the exercises. Results obtained from the
experiment are promising. The affective states recognized include the most
common in novice programmers according to the study of Bosch, D'Mello \& Mills
\cite{bixler2013detecting}: boredom, engagement/flow, confusion and frustration. For this experiment
four binary classifiers were compared: J-48 decision trees, k-nearest neighbour
classifier, feed forward neural network, and a naïve Bayes. The output of each
classifier determined if the student experienced the affective state during the
exercise. The proposed method could be used in an ensemble with other sensor
channels to improve the non-invasive recognition of the students’ affective
states when they are working on a programming task. In order to determine what
affective states a student was experiencing, an Experience Sampling Method (ESM)
was used by Csikszentmihalyi \& Larson \cite{kubey1996experience}. After the students successfully
solve a programming exercise, they are presented with an ESM survey that asks
what they were feeling during their solving of the exercise.

>>>>

When designing instruction, an objective is to assign learning activities
that result challenging to students, but not much as to frustrate them. For
this, experienced instructors gauge a student’s affective state and then assign
an activity with the appropriate level of difficulty. However, recognition of 
emotions is a difficult task, even
for humans. Instructors use their social intelligence often to recognize
students’ affective states. In class a good instructor habitually reads the
faces of students in the classroom to see if they are confused, bored or
engaged; then decides what to do next. Interactive Learning Environments (ILEs)
can be enhanced if they can offer an
automatic perception of emotions. The recognition and simulation of human
affects are becoming important fields of study, as many researchers have
demonstrated affect-aware computers can provide better performance in assisting
humans \cite{picard2001toward}. When ILEs embrace these methodologies and techniques a
problem arises. In order to perform recognition of the users' affective states,
sensors must be used to gather data. Some sensors rely on physiological readings
and must be in physical contact with students. These sensors can be considered
intrusive or invasive, and can disrupt the student’s learning experience 
\cite{zhai2008stress,sidney2005integrating,arroyo2009emotion}.
Other sensors such as video cameras can also be considered invasive, since they always transmit the
identity, appearance and behaviour along with emotional data \cite{picard2001toward}.

In this work, a method for the recognition of affective states through keystroke
and mouse dynamics is proposed. The hypothesis is that students vary the dynamic
of keystrokes according to their affective state when programming.  The method
has three main components: a Javascript library to capture keyboard and mouse
dynamics from a browser-based editor, a pre-processing step whose output is a
feature vector that is sent to the third component, a classification algorithm.
An experiment was conducted where students solved a series of programming
exercises using the web based learning environment. Using only the data from the
student’s keyboard and mouse dynamics six affective states where recognized for
each of the attempts at solving the exercises. Results obtained from the
experiment are promising. The affective states recognized include the most
common in novice programmers according to the study of Bosch, D'Mello \& Mills
\cite{bixler2013detecting}: boredom, engagement/flow, confusion and frustration. For this experiment
four binary classifiers were compared: J-48 decision trees, k-nearest neighbour
classifier, feed forward neural network, and a naïve Bayes. The output of each
classifier determined if the student experienced the affective state during the
exercise. The proposed method could be used in an ensemble with other sensor
channels to improve the non-invasive recognition of the students’ affective
states when they are working on a programming task. In order to determine what
affective states a student was experiencing, an Experience Sampling Method (ESM)
was used by Csikszentmihalyi \& Larson \cite{kubey1996experience}. After the students successfully
solve a programming exercise, they are presented with an ESM survey that asks
what they were feeling during their solving of the exercise.
